The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-10-06 thesis/extra_density.tex
full draft of appendicies, improved formatting

In this section, we describe the details of J+24's likelihoods. We then
test the reliability of the derived density profiles for Scl and UMi,
comparing methodologies, samples, and to literature. Finally, we present
a comparison of J+24's samples to a non-parametric Bayesian density
profile, finding similar results in the inner regions. Past a ``limiting
radius'', J+24's density profiles become susceptible to the likelihood
specification. However, all methods agree up to the limiting radii. In
conclusion, the density profiles from J+24 are robust except in the very
outer regions.

@@

To create a high-quality sample, J+24 select stars initially from Gaia
within a 2--4 degree circular region centred on the dwarf satisfying:

\begin{itemize}
\tightlist
\item
  Solved astrometry, magnitude, and colour.
\item
  Renormalized unit weight error, \({\rm ruwe} \leq 1.3\). Ensures high
  quality astrometry. \texttt{ruwe} is a measure of the excess
  astrometric noise on fitting a consistent parallax-proper motion
  solution \citep[see][]{lindegren+2021}.
\item
  3\(\sigma\) consistency of measured parallax with dwarf's distance,
  with \citet{lindegren+2021}'s zero-point correction. Note that the
  typical satellite parallax is very small.
\item
  Absolute proper motions, \(\mu_{\alpha*}\), \(\mu_\delta\), less than
  10\(\,{\rm mas\ yr^{-1}}\). This corresponds to tangental velocities
  of \(\gtrsim 500\) km/s at distances larger than 10 kpc.
\item
  Corrected colour excess is within 3\(\sigma\) of the expected
  distribution from \citet{riello+2021}. Removes stars with unreliable
  photometry.
\item
  De-reddened \(G\) magnitude is between
  \(22 > G > G_{\rm TRGB} - 5\sigma_{\rm DM}\). Removes very faint stars
  and stars significantly brighter than the tip of the red giant branch
  (TRGB) magnitude plus the distance modulus uncertainty
  \(\sigma_{\rm DM}\).
\item
  De-reddened colour is between \(-0.5 < {\rm BP - RP} <  2.5\). Removes
  stars substantially outside the expected CMD.
\end{itemize}

Photometry is dereddened with \citet{schlegel+finkbeiner+davis1998}
extinction maps.

J+24 define likelihoods \({\cal L}\) measuring consistency with the MW
stellar background (\({\cal L}_{\rm bg}\)) or the satellite galaxy
(\({\cal L}_{\rm sat}\)). The total likelihoods are the product of a
spatial, PM, and CMD term: \begin{equation}{
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}.
}\end{equation}

Each likelihood is normalized over their respective 2D parameter space
for both the satellite. To control the relative frequency of member and
background stars, \(f_{\rm sat}\) representing the fraction of member
stars in the field. The total likelihood for any star in this model is
the \(f_{\rm sat}\)-weighted sum of the foreground and background,
\begin{equation}\protect\phantomsection\label{eq:Ltot}{
{\cal L}_{\rm tot} = f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}.
}\end{equation} Finally, the probability that any star belongs to the
satellite is then given by \begin{equation}{
P_{\rm sat} = 
\frac{f_{\rm sat}\,{\cal L}_{\rm sat}}{{\cal L}_{\rm tot}}
= \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}}.
}\end{equation}

For the satellite's spatial likelihood, J+24 consider both one-component
and a two-component density models. The one component model is
constructed as a single exponential profile ( surface density
\(\Sigma \propto e^{R_{\rm ell} / R_s}\)), with scale radius \(R_s\).
\(R_s\) is fixed to the equivalent \(R_h\) value from
\citet{munoz+2018}'s Sérsic fit. Additionally, structural uncertainties
(for position angle, ellipticity, and scale radius) are sampled over to
construct the final likelihood map. The two-component model instead adds
a second exponential,
\(\Sigma_\star \propto e^{-R/R_s} + B\,e^{-R/R_{\rm outer}}\). The inner
scale radius is fixed, and the outer scale radius and magnitude of the
second component \(R_{\rm outer}\), \(B\) are free parameters.
Structural property uncertainties are not included in the two-component
model.

The PM likelihood is a bivariate gaussian with variance and covariance
equal to each star's proper motions. J+24 assume the stellar PM errors
are the main source of uncertainty.

The satellite's CMD likelihood is based on a Padova isochrone
\citep{girardi+2002}. The isochrone has a matching metallicity and 12
Gyr age (except 2 Gyr is used for Fornax). The isochrone is given a
Gaussian colour width of 0.1 mag plus the \emph{Gaia} colour uncertainty
at a given magnitude. The horizontal branch is modelled as a constant
magnitude extending blue of the CMD (mean magnitude of -2.2, 12 Gyr HB
stars and a 0.1 mag width plus the mean colour error). A likelihood map
is constructed by sampling the distance modulus in addition to the CMD
width, taking the maximum of RGB and HB likelihoods.

The background likelihoods are instead empirically constructed. Stars
stars outside of 5\(R_h\) passing the quality cuts estimate the
background density in PM and CMD space. The density is a sum of
bivariate gaussians with variances based on Gaia uncertainties (and
covariance for proper motions). The spatial background likelihood is
assumed to be uniform over the field.

J+24 derive \(\mu_{\alpha*}\), \(\mu_\delta\), \(f_{\rm sat}\) (and
\(B\), \(R_{\rm outer}\) for two-component) through an MCMC simulation
with likelihood from Eq.~\ref{eq:Ltot}. Priors are only weakly
informative. The proper motion single component prior is same as
\citet{MV2020a}: a normal distribution with mean 0 and standard
deviation \(100\ \kms\). If using the 2-component model, the prior is
instead a uniform distribution spanning 5\(\sigma\) of single component
case w/ systematic uncertainties. \(f_{\rm sat}\) (and \(B\)) has a
uniform prior from 0 to 1. \(R_{\rm outer}\) has a uniform prior only
restricting \(R_{\rm outer} > R_s\). The mode of each parameter from the
MCMC are then reported and used to calculate the final \(P_{\rm sat}\)
values.

\section{Alternative sample selection}\label{sec:simple_selection}

As J+24's probabilistic model may systematically bias the density
profile (see discussion below), we consider samples selected both from
\emph{Gaia} and external surveys using simple, absolute selection cuts.

For \emph{Gaia} data, we filter stars by four simple selection criteria.
We require \texttt{ruwe} \textless{} 1.3 as above, selecting only stars
with high-quality astrometry. We then require the star's parallax to be
\(3\sigma\)-consistent with the dwarf's distance. Next, we select stars
with proper motions with \(1\masyr\)\hspace{0pt} radius of the galaxy's
measured value. Finally, we filter stars to lie on the dwarf's CMD in
(\(G_{\rm BP} - G_{\rm RP}\),\(G\)), within the vertices defined in
Table~\ref{tbl:colour_cuts}. This polygon is derived based on the
colour-magnitude diagram of stars in the centre of the dwarf galaxy,
where the dwarf dominates the stellar density.

In addition to \emph{Gaia}, we select stars in the Scl field from the
deeper, photometric DELVE survey \citep{drlica-wagner+2022}. We select
sources DELVE data release 2 within an ellipse of radius 150 arcminutes.
We then only include sources which are categorized as likely stars and
sources with reliable (flags \textless=4) \(g\) and \(r\) magnitudes. We
then filter by the \(g\) and \(r\) magnitudes, selecting the empirical
CMD cut using the vertices in Table~\ref{tbl:colour_cuts}.

Ursa Minor is within the Ultraviolet Near-Infrared Optical Northern
Survey \citep[UNIONS][]{gwyn+2025}. We select sources within an ellipse
of radius 230 arminutes. We keep sources with no \texttt{FLAGS\_CFIS}
set and the \texttt{s21} and \texttt{s31} both \(<3\) (i.e.~not extended
sources). We then use the polygon bounded by the vertices in
Table~\ref{tbl:colour_cuts} to select the observed red giant branch,
horizontal branch, and main sequence of the dwarf.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/extra_cmd_selection.png}}
\caption[CMD cuts]{Colour-magnitude diagram cuts used in samples in this
section. Orange points are selected, black are non-selected stars within
\(1R_h\), and grey stars are for all stars. \textbf{todo, only plot
black/grey and cmd outline\ldots{}}}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/delve_unions_tangent.pdf}}
\caption[DELVE and UNIONS spatial distribution of stars]{The
distribution of DELVE and UNIONS selected stars in Scl and UMI}
\end{figure}

\section{\texorpdfstring{Uncertainties and possible biases in
\emph{Gaia}-derived density
profiles}{Uncertainties and possible biases in Gaia-derived density profiles}}\label{sec:density_extra}

In this section, we discuss additional tests and verification of the
derived density profiles. In particular, we check that alternative
methodologies do not substantially affect the density profile. In all
cases, the density profiles appear to have excellent convergence out to
\(\log R_{\rm ell} / {\rm arcmin} \approx 1.8\), about the distance
where the background dominates.

\textbf{Spatial likelihood.} J+24's algorithm was designed specifically
to detect the presence of a density excess in satellites and identify
individual stars for spectroscopic follow-up. We are instead interested
in constraining a dwarf galaxy's stellar structure. As such, one
potential problem with using J+24's candidate members is that the
stellar density model is either a single or double exponential. If the
(double) exponential model does not adequately describe a dwarf galaxy,
then J+24's derived density profiles may be systematically biased
towards the assumed (double) exponential profile.

As an example, in the lower panels of Fig.~\ref{fig:scl_density_extras};
Fig.~\ref{fig:umi_density_extras}, the density profile derived from the
fiducial (J+24) sample appears to be well-constrained at densities far
below the CMD+PM or \texttt{simple} model's background density. In
addition, the fiducial and 1-component density profiles deviate, despite
being derived identically besides the spatial likelihood assumption.

These stars are likely selecting stars from the statistical MW
background consistent either galaxy's CMD and PMs, recovering the
assumed density profile. As a result, the reliability of these density
profiles below the CMD+PM background may be questionable. A more robust
analysis, removing this particular density assumption, would be required
to more appropriately represent the knowledge of the density profile as
the background begins to dominate.

We revisit possible biases in spatial likelihood in
Section~\ref{sec:mcmc_hists}.

\textbf{Completeness}. \emph{Gaia} shows high but imperfect
completeness, particularly showing limitations in crowded fields and for
faint sources (\(G\gtrapprox20\)). In \citet{fabricius+2021}, for the
lowest density globular clusters, the completeness down to
\(G\approx 20\) is \(\sim 80\%\). As the lowest density globular
clusters are still far denser than dwarf galaxies, so the completeness
is likely higher for classical dwarfs. \emph{Gaia} is likely more
incomplete in terms of BP-RP photometry (as we use here) and for stars
with small pairwise separations (\(\lesssim 1.4''\)). The influence of
these details is unclear. Nevertheless, \emph{Gaia} should be more
complete for brighter stars.

We find little evidence for magnitude-dependent biases in the density
profiles. In
Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras} we show
density profiles derived using only stars brighter than the median
magnitude. We find no substantive differences compared to our fiducial
density profile. If \emph{Gaia}'s incompleteness is inhomogeneous, then
these variations do not appear to affect the density profiles.

If \emph{Gaia} has systematic density biases, the density profile should
vary among different facilities. We show the density profiles derived
from deep, ground-based UNIONS and DEVLE photometry (lower panels of
Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras}). For
both Scl and UMi, these alternative surveys result in near-identical
density profiles. Variations between density profiles derived from these
surveys appear to also be with uncertainties.

\textbf{Structural uncertainties}. The assumed structural parameters of
a dwarf may be misrepresentative or vary with radius. J+24 derive
probabilities where the outer stellar component is circular instead of
elliptical. Using the circular 2-component probabilities and binning in
circular radius, we find little difference in the derived density
profiles (the \texttt{circ} points in
Figs.~\ref{fig:scl_density_extras}, \ref{fig:umi_density_extras}). Thus,
our conclusions are robust to extreme variations in the ellipticity.
Reasonable changes in other structural parameters (centre, position
angle) are unlikely to have a more substantial effect.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/density_methods_extra.pdf}}
\caption[Sculptor density methodology comparison]{Density profiles for
various assumptions for Sculptor. \texttt{2-exp} is the fiducial
double-exponential-likelihood J+24 sample, \texttt{1-exp} instead is a
one-component exponential spatial likelihood, and \texttt{simple} uses
the simple position-independent selection criteria \texttt{circ} is a
2-component bayesian model assuming circular radii, simple is the series
of simple cuts described, bright is the sample of the brightest half of
stars (scaled by 2).}\label{fig:scl_density_extras}
\end{figure}

@@

To address the concerns discussed above, we consider here a
non-parametric model to fit the density in each bin.

\begin{table*}[t]
\centering
\caption[Properties of probabilistically-derived density profiles]{For each classical dwarf, we have: the BG-limited radius $R_{\rm limit}$, where the density of stars is no longer reliably derived, the approximate number of member stars in Gaia, and the derived Sérsic indices from the density profiles. }
\label{tbl:mcmc_props}
\begin{tabular}{lllll}
\toprule
Galaxy & $R_{\rm limit} / R_h$ & $R_{\rm limit} / '$ & num stars & $n$\\
\midrule
Fornax & 5.25 & 79.1 & 23154 & $0.794^{+0.034}_{-0.018}$\\
Sculptor & 6.39 & 64.1 & 7024 & $1.281^{+0.04}_{-0.04}$\\
Leo I & 4.23 & 13.5 & 1242 & $0.719^{+0.059}_{-0.053}$\\
Ursa Minor & 6.42 & 86.4 & 2314 & $1.425^{+0.10}_{-0.093}$\\
Leo II & 3.63 & 8.76 & 347 & $0.707^{+0.11}_{-0.092}$\\
Carina & 4.16 & 33.3 & 2389 & $1.025^{+0.08}_{-0.076}$\\
Draco & 3.59 & 27.3 & 1781 & $1.106^{+0.084}_{-0.078}$\\
Canes Venatici I & 1.95 & 12.5 & 156 & $1.18^{+0.36}_{-0.28}$\\
Sextans I & 3.42 & 67.9 & 1830 & $1.136^{+0.10}_{-0.09}$\\
Crater II & 1.93 & 39.0 & 507 & $0.535^{+0.12}_{-0.09}$\\
\bottomrule
\end{tabular}
\end{table*}

@@

This model extends the J+24 framework with two notable differences. (1)
The systemic proper motions are held fixed. This allows for a much more
efficient calculation of the likelihoods. (2) The spatial likelihood for
the satellite is a piecewise constant density function.While this model
theoretically has many more parameters to fit (one for each bin), each
parameter is independently estimated based on only the stars in that
specific bin.

We segregate the data into radial bins and independently fit the model
to only stars in each bin. This fits many small models instead of a
global large model. In detail, this means that, in each bin, the
normalized density of the background and satellite are \emph{equal}. The
only fit parameter is the \(f_{\rm sat}\) in the particular bin.
Therefore, the likelihood in each bin is the same as above except
\({\cal L}_{\rm space} = 1\) for the foreground and background.

In detail, we parameterize \(f_{\rm sat}\) in terms of the log-relative
satellite density. The model construction is then \begin{equation}{
\begin{align}
\theta_i &\equiv \log_{10}({\Sigma_{\rm sat}}/{\Sigma_{\rm bg}}) \\
f_{\rm sat} &= \frac{10^{\theta_i}}{1 + 10^{\theta_i}} \\
\theta_i &\sim {\rm Uniform}(-12, 6) \\
{\cal L}_{\rm tot} &= f_{\rm sat}{\cal L}_{\rm CMD, PM} + (1-f_{\rm sat}){\cal L}_{\rm CMD, PM} \\
\end{align}
}\end{equation}

For each satellite, we hold the structural parameters fixed to those in
J+24, create bins which are the larger of the bin containing 20 stars or
0.05dex in \(\log R_{\rm ell}\). We then sample each chain 1000 steps
for 48 walkers using Turing.jl and the No-U Turns Sampler.

Based on the \(f_{\rm sat}\) values in each bin, the number of members
is just the number of stars in the bin times \(f_{\rm sat}\) for the
bin, allowing the direct derivation of the density profiles from this
model.

We have also briefly explored a model which solves for \(f_{\rm sat}\)
for random realizations of the global structural uncertainties. These do
not appear to influence the results substantially. Another possible
inconsistency is that the half-light radii are estimated from a
different sample. We do briefly rederive structural parameters for the
sample using Sérsic fits (although with a fixed origin).

@@

Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived, MCMC
histogram density profiles compared to J+24. In general, both
methodologies agree well. However, J+24 tend to systematically
overestimate faint densities and confidently derive densities where the
MCMC model fails. Because J+24 use only one or two components across the
entire dwarf galaxy, the faint regions become sensitive to the spatial
likelihood. Antlia II is an extreme case, where the satellite is hidden
behind a large number of foreground stars. We derive a much more
poorly-constrained and fainter total density profile than J+24, and the
divergence between the MCMC method here and J+24's profile likely shows
that \emph{Gaia} observations are insufficient to properly constrain
Antia II's density profile. This is likely due to uncertainties in the
background density of satellite-like stars in Antlia II.

Once the true density of stars drops below the background of
satellite-like stars, J+24's method will occasionally select consistent
background stars with a density directly dependent on the spatial
likelihood. The uncertainties in the outer regions are likely
underestimated.

To properly compare density profiles before background-limiting effects
become important, we only calculate our profiles in the main text out to
the radii in Table~\ref{tbl:mcmc_props}. We derive these
``background-limited'' radii based on the outermost derived density in
the MCMC histogram with an uncertainty lower than 1 dex. This typically
corresponds to the empirical background from the CMD+PM samples in the
main text.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/mcmc_histograms.png}}
\caption[Probabilistic density profiles]{A comparison between the MCMC
histogram method and J+24. The MCMC samples in each bin are black
transparent dots, and the J+24 derived density profiles (with their
\(P_{\rm sat} < 0.2\)) are orange solid dots.}\label{fig:mcmc_hists}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/mcmc_histograms2.png}}
\caption[Probabilistic density profiles continued]{A continuation of
Fig.~\ref{fig:mcmc_hists}.}\label{fig:mcmc_hists2}
\end{figure}

@@

We finally compare our density profiles against a sample of
literature-derived density profiles in
Figs.~\ref{fig:scl_lit_profiles}, \ref{fig:umi_lit_profiles}. The
J+24-derived density profiles are representative of state-of-the-art
density profiles. A few photometric surveys may extend to slightly
fainter densities, but typically with large uncertanties. Deviations
between sources are slight despite the range of methods across time. All
density profiles extending into the outskirts of Scl and UMi show a
similar overdensity to the ones we find. The extended stellar profiles
of Scl and UMi appear to be a robust result across the literature.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_literatre_profiles.pdf}}
\caption[Sculptor literature density profiles]{A comparison of the Scl
derived density profile and historical works. Unlike most profiles in
this thesis, this density profile is plotted with respect to the
semi-major elliptical radius (\(a = R_{\rm ell} / \sqrt{1-{\rm ell}}\)).
The solid black line is a 2D exponential with corresponding scale radius
to Scl's half-light radius, and the residuals in the bottom panel are
with respect to this profile. \textbf{remove delve, umi title\ldots,,
combine??}}\label{fig:scl_lit_profiles}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_literature_profiles.pdf}}
\caption[Ursa Minor literature density profiles]{Similar to
Fig.~\ref{fig:scl_lit_profiles} except for Ursa Minor. The references
are (in order) \citet{sato+2025}; \citet{palma+2003};
\citet{martinez-delgado+2001}; \citet{kleyna+1998}; \citet{IH1995}; and
\citet{Hodge1964}.}
\end{figure}

>>>>

In this section, we detail the J+24 membership algorithm and test the
resulting density profiles for Scl and UMi. We compare profiles derived
with alternative methodologies and literature results. We also present a
non-parametric Bayesian density profile, which reproduces J+24's results
in the inner regions but diverges when the background dominates. In
summary, we find the density profiles are robust up to the
background-limited radius derived here.

@@

To create a high-quality sample, J+24 select stars initially from Gaia
within a 2--4 degree region around each satellite with high-quality
astrometry, reliable photometry, and consistent parallaxes and broadly
consistent proper motions and colours. Stars are removed if they have
excess astrometric noise \citep[\({\rm ruwe} > 1.3\),
see][]{lindegren+2021}, colour excess \(3\sigma\) outside of
expectations \citep[from][]{riello+2021}, proper motions magnitudes
\(>10\,\masyr\) in \(\mu_{\alpha*}\) or \(\mu_\delta\), magnitudes
brighter than the tip of the red giant branch or fainter than \(G=22\),
or colours outside \(-0.5 < G_{\rm BP} - G_{\rm RP} <  2.5\). Photometry
is dereddened with \citet{schlegel+finkbeiner+davis1998} extinction
maps.

J+24 construct satellite likelihoods in spatial, proper motion, and CMD
space following expected satellite properties. J+24 model the spatial
likelihood as either a one or two-component exponential
(Eq.~\ref{eq:exponential_law}). The structural parameters of the inner
component are fixed, and marginalized over if one-component. The outer
profile scale radius and normalization are free parameters.

The PM likelihood is a bivariate gaussian with variance and covariance
equal to each star's proper motions.

J+24 model the CMD as a Padova isochrone \citep{girardi+2002}, with
matching metallicity, 12 Gyr age (or 2 Gyr for Fornax), and Gaussian
width of 0.1 mag plus the \emph{Gaia} colour uncertainty at each
magnitude. The horizontal branch is modelled as a
constant-magnitude\footnote{Specifically, the mean magnitude of the 12
  Gyr, \({\rm [Fe/H]}=-2.2\), Padova isochrone's horizontal branch.}
sequence extending blue of the CMD with the same width as the RGB. The
CMD likelihoods are marginalized over the distance modulus and take the
maximum of the RGB and horizontal branch likelihoods.

The background likelihoods are determined empirically as a kernel
density estimates from stars outside 5\(R_h\) in PM and CMD space. The
spatial background likelihood is uniform.

J+24 derive the distributions of parameters (proper motions, satellite
fraction \(f_{\rm sat}\), and second spatial component if included)
through Monte Carlo Markov chain sampling with broad or
weakly-informative priors. The posterior modes are used to calculate the
final \(P_{\rm sat}\) values.

\section{\texorpdfstring{Possible biases in \emph{Gaia}-derived density
profiles}{Possible biases in Gaia-derived density profiles}}\label{sec:density_extra}

In this section, we test how assumptions in J+24's algorithm or biases
in \emph{Gaia} may affect density profiles. In all cases, density
profiles converge until \(R_{\rm ell} \approx 60\,{\rm arcmin}\), where
background contamination likely dominates.

\textbf{Spatial likelihood.} As J+24 include a spatial likelihood, our
density profiles risks ``double fitting''---the spatial distribution of
stars informs the membership catalogue from which density profiles are
derived. If the assumed spatial likelihood is misrepresentative, the
derived density profiles may bias towards the assumed likelihood.
Fig.~\ref{fig:scl_umi_density_extras} illustrates that selecting the
one- or two-component spatial models affects the outer extent of the
dwarf galaxy. Promisingly, the density excess persists when assuming a
one-component exponential profile. We revisit this concern using
spatially independent methods below.

\textbf{Structural uncertainties}. The assumed structural parameters of
a dwarf (centre, position angle, ellipticity, scale radius) may be
biased or vary radially. As a test, we use a membership sample J+24
derive assuming the outer component is circular. Even when binning in
circular radius, the density profiles remain unchanged (the
\texttt{circ} points in Fig.~\ref{fig:scl_umi_density_extras}). Thus,
our conclusions appear to be robust to an extreme shift in ellipticity.
Reasonable changes in other structural parameters are unlikely to have a
stronger effects.

\textbf{Completeness}. \emph{Gaia} appears to have high but imperfect
completeness, more limited for crowded fields, faint sources
(\(G\gtrsim20\)), and BP-RP magnitudes. In \citet{fabricius+2021}, the
completeness down to \(G\approx 20\) is \(\sim 80\%\) for low-density
globular clusters. As dwarfs are even less dense, their completeness is
likely higher. To test for magnitude-dependent biases, we show density
profiles derived for only the brighter half of member stars in
Fig.~\ref{fig:scl_umi_density_extras}. We find no substantive change
compared to our fiducial density profile. If \emph{Gaia}'s
incompleteness is inhomogeneous, then these variations likely do not
affect our density profiles.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/density_methods_extra.pdf}}
\caption[Sculptor density methodology comparison]{Density profiles for
various assumptions for Sculptor. \texttt{2-exp} is the fiducial
double-exponential-likelihood J+24 sample, \texttt{1-exp} instead is a
one-component exponential spatial likelihood, \texttt{simple} uses the
simple position-independent selection criteria, \texttt{circ} is a
2-component bayesian model assuming circular radii, \texttt{bright} only
includes stars brighter than the median magnitude, and \texttt{DELVE} or
\texttt{UNIONS} use photometry from external surveys with background
subtraction. In all cases, the density profiles are nearly identical
until the background-limited regime (grey shaded
region).}\label{fig:scl_umi_density_extras}
\end{figure}

\section{Comparison against alternate
samples}\label{sec:simple_selection}

Next, to compare against simpler methods, we create samples using
absolute filters in the CMD, PM space. We also compare against similarly
selected samples from deep photometry-only data.

For the \texttt{simple} \emph{Gaia} samples, we require high-quality
astrometry (\texttt{ruwe} \textless{} 1.3), parallaxes
\(3\sigma\)-consistent with the dwarf's distance, proper motions with
\(1\masyr\) radius of the galaxy's value, and within the dwarf's
empirical CMD (Fig.~\ref{fig:extra_cmd}).

We also determine density profiles from the deeper photometric surveys.
For Scl, we use DELVE DR2 survey \citep{drlica-wagner+2022}. We select
sources within an ellipse of radius 150 arcminutes, categorized as
likely stars, with reliable \(g\) and \(r\) magnitudes (associated flags
\(\leq 4\)), and within Scl's observed CMD (Fig.~\ref{fig:extra_cmd}).
Ursa Minor is within the Ultraviolet Near-Infrared Optical Northern
Survey \citep[UNIONS,][]{gwyn+2025}. We select sources within an ellipse
of radius 230 arminutes, with no \texttt{FLAGS\_CFIS} set, the
\texttt{s21} and \texttt{s31} both \(<3\) (i.e.~not extended sources),
and within the CMD cut in Fig.~\ref{fig:extra_cmd}.
Fig.~\ref{fig:delve_unions_tangent} shows the resulting spatial
distribution for both samples. The only remarkable structure is Muñoz 1
nearby to Ursa Minor.

Finally, Fig.~\ref{fig:scl_umi_density_extras} compares the density
profiles from each sample. All samples agree until the ``limiting
radius'', where the \texttt{simple} and DELVE/UNIONS samples reach their
respective backgrounds. The very outskirts of J+24's density profiles
may also extend below the background of apparent member stars,
complicating the reliability of the outer density profiles.

\begin{figure}
\centering
\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{figures/extra_cmd_selection.png}
\caption[Colour-Magnitude sample selection]{Colour-magnitude diagram
cuts used in samples in this section. The likely-member stars in the
inner \(1R_h\) of each dwarf (black points) are used as a guide to
determine the CMD cuts (orange polygons). Light grey points instead show
the full distribution of stars across the field.}\label{fig:extra_cmd}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/delve_unions_tangent.pdf}}
\caption[DELVE and UNIONS spatial distribution of stars]{The
distribution of DELVE and UNIONS selected stars in Scl and UMi as grey
points overdrawn with isophotes. The clump to the lower right of UMi is
Muñoz 1.}\label{fig:delve_unions_tangent}
\end{figure}

@@

To address the concerns discussed above, we consider here a
non-parametric model to fit the density in each bin. We demonstrate that
the non-parametric model fails to find evidence for stellar density
where J+24 still detects members. Notably, our density profiles diverge
substantially from J+24 for Antlia II. We suggest these differences
arise from background contamination, motivating our ``limiting radii''
representing where densities may become unreliable.

@@

As a non-parametric but similar model to J+24, we consider a piecewise
constant spatial likelihood. In this model, the stars are divided into
radial bins which are then considered independently. Each bin has only
one free parameter, the fraction of satellite stars in that bin,
\(f_{\rm sat}\). This model thus directly derives the density profile
from the data in a single step. If a bin contains insufficient
information to estimate a precise satellite density, than the posterior
\(f_{\rm sat}\) should reflect this uncertainty.

We parameterize \(f_{\rm sat}\) in terms of a log-relative density,
\(\theta\) \begin{equation}{
\begin{split}
\theta &\equiv \log_{10}({\Sigma_{\rm sat}}/{\Sigma_{\rm bg}}) \\
f_{\rm sat} &= \frac{10^{\theta}}{1 + 10^{\theta}}.
\end{split}
}\end{equation} We then adopt a broad uniform prior on \(\theta\) from
-12 to 6. The CMD and PM likelihoods are unchanged from J+24, except the
systemic PM is fixed for efficiency.

To bin stars, we hold fixed J+24's structural parameters and create bins
of the wider of the width \(\Delta \log R=0.05\) or containing the next
20 stars. We then derive posterior \(f_{\rm sat}\) distributions with
MCMC (48 walkers of 1000 steps each, and using the No-U-Turn sampler as
implemented in Turing.jl). The final density profile is directly derived
from \(f_{\rm sat}\) and the number of stars in each bin.

@@

\begin{table*}[t]
\centering
\caption[The limiting radii of Gaia-derived density profiles]{For each classical dwarf, the limiting radius $R_{\rm limit}$ in units of $R_h$ and arcminutes. $R_{\rm limit}$ represents where there no longer appears to be evidence of stars in Emph(Str(Gaia)) using the nonparametric MCMC density profiles. }
\label{tbl:mcmc_props}
\begin{tabular}{lll}
\toprule
Galaxy & $R_{\rm limit} / R_h$ & $R_{\rm limit} / '$\\
\midrule
Fornax & 5.25 & 79.1\\
Sculptor & 6.39 & 64.1\\
Leo I & 4.23 & 13.5\\
Ursa Minor & 6.42 & 86.4\\
Leo II & 3.63 & 8.76\\
Carina & 4.16 & 33.3\\
Draco & 3.59 & 27.3\\
Canes Venatici I & 1.95 & 12.5\\
Sextans I & 3.42 & 67.9\\
Crater II & 1.93 & 39.0\\
\bottomrule
\end{tabular}
\end{table*}

Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived, MCMC
density profiles as compared to J+24. In general, both methodologies are
consistent. However, J+24 tend to systematically overestimate faint
densities and confidently derive densities where the MCMC model fails to
derive a density estimate.

In the outskirts of satellites, more background / foreground stars may
have consistent PM and CMD properties than satellite members. Improperly
estimating the satellite's density in this ``background-limited'' regime
likely affects the inferred density of members. If the satellite's
density is severely overestimated, then a J+24-like sample may select
many additional background stars. The density of candidates would then
be biased towards the assumed density. As J+24 only use a one or
two-component density profile across the entire dwarf, the density
profile is likely biased where the CMD+PM background dominates. The
nonparametric MCMC model instead does not assume a local density, so
should better represent the underlying satellite density.

Antlia II represents an extreme example---J+24 and the non-parametric
density model systematically disagree across the entire galaxy.
Additionally, the piecewise model derives densities with larger
uncertainties and over a smaller range than J+24. The methodological
divergence here likely arises due to an extraordinary
background/foreground of MW stars. \emph{Gaia} data alone may be unable
to properly constrain the density of such background-contaminated
objects.

To properly compare density profiles before background-limiting effects
become important, we only calculate our profiles in the main text out to
the radii in Table~\ref{tbl:mcmc_props}. We derive these
``background-limited'' radii based on the outermost derived density in
the MCMC non-parametric model with an uncertainty lower than 1 dex. This
closely corresponds to the background density from the ``CMD+PM''
samples in the main text (see Fig.~\ref{fig:scl_observed_profiles}).

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/mcmc_histograms.png}
\caption[Probabilistic density profiles]{A comparison between the MCMC
histogram method and J+24. The MCMC samples in each bin are black
transparent dots (with added jitter), and the J+24 derived density
profiles (i.e., \(P_{\rm sat} < 0.2\)) are orange solid
dots.}\label{fig:mcmc_hists}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/mcmc_histograms2.png}
\caption[Probabilistic density profiles
continued]{Fig.~\ref{fig:mcmc_hists} continued.}\label{fig:mcmc_hists2}
\end{figure}

@@

Finally, we compare our density profiles against a sample of
literature-derived density profiles in
Figs.~\ref{fig:scl_lit_profiles}, \ref{fig:umi_lit_profiles}. Deviations
between different profiles are slight, despite the range of methods
across time, and all density profiles extending into the outskirts of
Scl and UMi show a similar overdensity to the ones we find. The extended
stellar profiles of Scl and UMi appear to be a robust result across the
literature.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_literatre_profiles.pdf}}
\caption[Sculptor literature density profiles]{A comparison of the Scl
derived density profile and historical works. Unlike most profiles in
this thesis, this density profile is plotted with respect to the
semi-major elliptical radius (\(a = R_{\rm ell} / \sqrt{1-{\rm ell}}\)).
The solid black line is a 2D exponential with corresponding scale radius
to Scl's half-light radius, and the residuals in the bottom panel are
with respect to this profile. References are (in order),
\citet{munoz+2018}; \citet{westfall+2006}; \citet{walcher+2003};
\citet{eskridge1988}; \citet{demers+krautter+kunkel1980}; and
\citet{hodge1961}.}\label{fig:scl_lit_profiles}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_literature_profiles.pdf}}
\caption[Ursa Minor literature density profiles]{Similar to
Fig.~\ref{fig:scl_lit_profiles} except for Ursa Minor. The references
are (in order) \citet[derived using their minor axis
profile]{sato+2025}; \citet{palma+2003}; \citet{martinez-delgado+2001};
\citet{kleyna+1998}; \citet{IH1995}; and
\citet{Hodge1964}.}\label{fig:umi_lit_profiles}
\end{figure}
